---
title: "Alternating Manifold Proximal Gradient Method"
author: "Shixiang Chen, Justin Huang, Benjamin Jochem, Lingzhou Xue, Hui Zou"
date: "`r Sys.Date()`"
output:
  pdf_document: default
---

\newcommand{\argmin}{\operatorname{arg\,min}}

```{r setup, include=FALSE}
library(kableExtra)
knitr::opts_chunk$set(echo = TRUE, python.reticulate=FALSE)
```

## Contents

- [Introduction](#introduction)
  - [Algorithm](#algorithm-description)
  - [Convergence](#convergence)
  - [Pseudocode](#pseudocode)
- [Installation](#installation)
- [Documentation](#documentation)
	- [R Usage](#r-usage)
	- [Python Usage](#python-usage)
	- [Arguments](#arguments)
	- [Values](#values)
- [Quick Start](#quick-start)
	- [R Quick Start](#r-example)
	- [Python Example](#python-example)
- [References](#references)

## Introduction

`sparsepca` and `amanpg` find sparse loadings in principal component analysis (PCA) via an alternating manifold proximal gradient method (A-ManPG). Seeking a sparse basis allows the leading principal components to be easier to interpret when modeling with high-dimensional data. PCA is modeled as a regularized regression problem under the elastic net constraints (linearized L1 and L2 norms) in order to induce sparsity. Due to the nonsmoothness and nonconvexity numerical difficulties, A-ManPG is implemented to guarantee convergence. 

The package provides a function for performing sparse PCA and a function for normalizing data.

The authors of A-ManPG are Shixiang Chen, Shiqian Ma, Lingzhou Xue, and Hui Zou. The Python and R packages are maintained by Justin Huang and Benjamin Jochem. A MATLAB implementation is maintained by Shixiang Chen.

#### Algorithm Description

The A-ManPG algorithm can be applied to solve the general manifold optimization problem

\begin{equation}
  \text{min}~F(A,B) := H(A,B)+f(A)+g(B)\text{ subject to (s.t.) }A\in\mathcal{M}_1, B\in\mathcal{M}_2
\end{equation}

where $H(A,B)$ is a smooth function of $A,B$ with a Lipschitz continuous gradient, $f(\cdot)$ and $g(\cdot)$ are lower semicontinuous (possibly nonsmooth) convex functions, and $\mathcal{M}_1,\mathcal{M}_2$ are two embedded submanifolds in the Euclidean space.

For sparse PCA, the following function definitions are used:

- $H(A,B)=\text{Tr}(B^TX^TXB)-2\text{Tr}(A^TX^TXB)$
- $f(A)\equiv 0$
- $g(B)=\lambda_2\sum_{j=1}^k\|B_j\|_2^2+\sum_{j=1}^k\lambda_{1,j}\|B_j\|_1$
- $\mathcal{M}_1=\text{St}(p,k)$
- $\mathcal{M}_2=\mathbb{R}^{p\times k}$ 

where $X$ is the $n\times p$ data matrix or $n\times n$ covariance matrix, $A$ is the scores and $B$ is the loadings, $k$ is the rank of the matrices (in other words, how many principal components are desired), $\lambda_1$ is the L1 norm penalty and $\lambda_2$ is the L2 norm penalty. Both the L1 and L2 norm are used as elastic net regularization to impose sparseness within the loadings. Note that a different L1 norm penalty is used for every principal component, and the algorithm operates differently when the L2 norm penalty is set to a large constant (`np.inf` or `Inf`).

The A-ManPG algorithm uses the following subproblems with an alternating updating scheme to solve sparse PCA, computed in a Gauss-Seidel manner for faster convergence.

\begin{align}
  D_k^A := \argmin_{D^A} \langle\nabla_A H(A_k,B_k),D^A\rangle + f(A^k + D^A) + \frac{1}{2t_1}\|D^A\|^2_F~\text{ s.t. }~D^A\in\text{T}_{A_k}\mathcal{M}_1 \\
  D_k^B := \argmin_{D^B} \langle\nabla_A H(A_{k+1},B_k),D^B\rangle + f(B^k + D^B) + \frac{1}{2t_2}\|D^B\|^2_F~\text{ s.t. }~D^B\in\text{T}_{B_k}\mathcal{M}_2
\end{align}

where $A_{k+1}$ is obtained via a retraction operation (in this case, polar decomposition), $t_1\leq L_A$, and $t_2\leq L_B$. $L_A$ and $L_B$ are the least upper bounds of the Lipschitz constants for $\nabla_A H(A,B)$ and $\nabla_B H(A,B)$, respectively. The subproblems are solved using an adaptive semismooth Newton method.

#### Convergence

Let $\epsilon$ represent a tolerance level to detect convergence. An *$\epsilon$-stationary point* is defined as a point $(A, B)$ with corresponding $D^A$ and $D^B$ that satisfy the following:

\begin{equation}
  \|D^A/t_1\|_F^2+\|D^B/t_2\|_F^2 \leq \epsilon^2
\end{equation}

The algorithm reaches an 
$\epsilon$-stationary point in at most

$$\frac{2(F(A_0,B_0)-F^*)}{ ((\gamma\bar{\alpha}_1t_1+\gamma\bar{\alpha}_2t_2)\epsilon^2)}$$

iterations, where:

- $(A_0,B_0)$ are the initial values
- $F^*$ is the lower bound of $F$, from the general manifold optimization problem
- $\bar{\alpha}_1$ and $\bar{\alpha}_2$ are positive constants

#### Pseudocode

The following describes the algorithm used for solving the general manifold optimization problem using A-ManPG. For solving sparse PCA, the algorithm is implemented with the aforementioned definitions.

```{r tidy=FALSE, eval=FALSE, highlight=FALSE}
Input initial point (A0,B0) and necessary parameters for the required problem 

for i=0,1,... do
  Solve the first subproblem for Da
  Set alpha = 1
  
  while F(Retr(alpha * Da),B) > F(A,B) - alpha * norm(Da)^2 / (2 * t1) do
    alpha = gamma * alpha
  end while
  
  Set A = Retr(alpha * Da)
  
  Solve the second subproblem for Db
  Set alpha = 1
  
  while F(A,Retr(alpha * Db)) > F(A,B) - alpha * norm(Db)^2 / (2 * t2) do
    alpha = gamma * alpha
  end while
  
  Set B = Retr(alpha * Db)
end for

Return A as the scores and B as the sparse loadings
```

## Installation

To install the R package, install `amanpg` directly from CRAN.
```{r eval=FALSE, highlight=FALSE}
install.packages("amanpg")
```

To install the Python package, use `pip` to obtain `sparsepca` from PyPI.
```{python eval=FALSE, highlight=FALSE}
pip3 install sparsepca
```

## Documentation

#### R Usage

```{r eval=FALSE, highlight=FALSE}
spca.amanpg(z, lambda1, lambda2, 
			f_palm = 1e5, x0 = NULL, y0 = NULL, k = 0, type = 0, 
			gamma = 0.5, maxiter = 1e4, tol = 1e-5, 
			normalize = TRUE, verbose = FALSE)
```

#### Python Usage

```{python eval=FALSE, highlight=FALSE}
spca(z, lambda1, lambda2, 
     x0=None, y0=None, k=0, gamma=0.5, type=0, 
     maxiter=1e4, tol=1e-5, f_palm=1e5,
	   normalize=True, verbose=False):
```

#### Arguments

| Name | Python Type | R Type | Description |
| --- | --- | --- | --- |
| `z` | numpy.ndarray | matrix | Either the data matrix or sample covariance matrix |
| `lambda1` | float list | numeric vector | List of parameters of length n for L1-norm penalty |
| `lambda2` | float or numpy.inf | numeric or Inf | L2-norm penalty term |
| `x0` | numpy.ndarray | matrix | Initial x-values for the gradient method, default value is the first n right singular vectors |
| `y0` | numpy.ndarray | matrix | Initial y-values for the gradient method, default value is the first n right singular vectors |
| `k` | int | int | Number of principal components desired, default is 0 (returns min(n-1, p) principal components) |
| `gamma` | float | numeric | Parameter to control how quickly the step size changes in each iteration, default is 0.5 |
| `type` | int | int | If 0, b is expcted to be a data matrix, and otherwise b is expected to be a covariance matrix; default is 0 |
| `maxiter` | int | int | Maximum number of iterations allowed in the gradient method, default is 1e4 |
| `tol` | float | numeric | Tolerance value required to indicate convergence (calculated as difference between iteration f-values), default is 1e-5 |
| `f_palm` | float | numeric | Upper bound for the F-value to reach convergence, default is 1e5 |
| `normalize` | bool | logical | Center and normalize rows to Euclidean length 1 if True, default is True |
| `verbose` | bool | logical | Function prints progress between iterations if True, default is False |e

#### Values

Python returns a dictionary with the following key-value pairs, while R returns a list with the following elements:

| Key | Python Value Type | R Value Type | Value |
| --- | --- | --- | --- |
| `loadings` | numpy.ndarray | matrix | Loadings of the sparse principal components |
| `f_manpg` | float | numeric | Final F-value |
| `x` | numpy.ndarray | matirx | Corresponding ndarray in subproblem to the loadings |
| `iter` | int | numeric | Total number of iterations executed |
| `sparsity` | float | numeric | Number of sparse loadings (`loadings == 0`) divided by number of all loadings |
| `time` | float | numeric | Execution time in seconds |

## Quick Start

Consider the two examples below for running sparse PCA on randomly-generated data: one using a finite $\lambda_2$, and the other using a large constant $\lambda_2$.

#### R Example

As with other libraries, begin by loading `amanpg` in R.

```{r}
library(amanpg)
```

Before proceeding, it is helpful to determine a few parameters. Let the rank of the sparse loadings matrix be $k=4$ (returning four principal components), the input data matrix be $n\times p$ where $n=1000$ and $p=500$, $\lambda_1$ be a $4\times 1$ "matrix" where $\lambda_{i,1}=0.1$, and $\lambda_2=1$.

```{r}
# parameter initialization
k <- 4
n <- 1000
p <- 500
lambda1 <- matrix(data=0.1, nrow=4, ncol=1)
lambda2 <- 1
```

For this example, the data matrix `z` is randomly generated from the normal distribution. Although it should be centered to mean 0 and normalized to Euclidean length 1, the function will automatically preprocess the input matrix when `normalize=TRUE`. 

```{r}
# data matrix generation
set.seed(10)
z <- matrix(rnorm(n * p), n, p)
```

```{r echo=FALSE}
kbl(head(as.data.frame(z))[, 1:10], caption="Generated Data Matrix", booktabs=T) %>%
  kable_styling(latex_options=c("striped", "hold_position", "scale_down"))
```

Alternatively, the data can be normalized beforehand and the parameter is set to `FALSE` in the function call. However, this example won't do so, but the output of normalize is displayed below.

```{r eval=FALSE}
# see the effects of normalize()
head(as.data.frame(normalize(z))[, 1:10])
```

```{r echo=FALSE}
kbl(head(as.data.frame(normalize(z))[, 1:10]), caption="Normalized Data Matrix", booktabs=T) %>%
  kable_styling(latex_options=c("striped", "hold_position", "scale_down"))
```

Now the function is called, passing through matrix `a`, `lambda1`, `lambda2`, and our desired rank `k`. The output is stored as a list in `fin_sprout`. Note that if a different initial point is desired, `x0` and `y0` should be modified, but the default value as the first $k$ right singular vectors is sufficient for this example.

If further printout is desired, set `verbose=TRUE` for progress updates (time, difference for convergence, $F$ value) per iteration.

```{r}
# function call
fin_sprout <- spca.amanpg(z, lambda1, lambda2, k=4)
print(paste(fin_sprout$iter, "iterations,", fin_sprout$sparsity, "sparsity,", fin_sprout$time))
```

The loadings can be viewed from `fin_sprout$loadings`. Note that many entries are set to zero as a result of the induced sparsity.

```{r eval=FALSE}
# View loadings
head(as.data.frame(fin_sprout$loadings))
```

```{r echo=FALSE}
kbl(head(fin_sprout$loadings), booktabs = T, caption = "Finite Lambda Loadings") %>%
  kable_styling(latex_options = c("striped", "repeat_header", "hold_position"))
```

The resulting scree plot looks like this.

```{r fig.align='center', fig.cap="Scree plots for the finite lambda case."}
pr.var <- (apply(fin_sprout$x, 2, sd))^2
pve <- pr.var / sum(pr.var)

par(mfrow=c(1,2))
plot(pve, 
     xlab="Sparse PC", 
     ylab="Proportion of Variance Explained", 
     ylim=c(0,1), 
     type="b")
plot(cumsum(pve), 
     xlab="Sparse PC", 
     ylab="Cumulative Proportion of Variance Explained", 
     ylim=c(0,1),
     type="b")
```

The resulting biplot, with the zero loadings filtered out, can be obtained using the following:

```{r fig.align='center', fig.cap="Biplot for the finite lambda case."}
y_sub = apply(fin_sprout$loadings, 1, function(row) all(row != 0))
loadings = fin_sprout$loadings[y_sub, ]
par(mfrow=c(1,1))
biplot(fin_sprout$x, loadings, xlab="PC 1", ylab="PC 2")
```

\newpage

Now consider an alternative situation where we set $\lambda_2$ to a large constant `Inf`. The algorithm changes by directly retracting `B` without using a while loop to determine an appropriate retraction step size and only iterating `A`. 

```{r}
# infinite lambda2
inf_sprout <- spca.amanpg(z, lambda1, lambda2=Inf, k=4) 
print(paste(inf_sprout$iter, "iterations,", inf_sprout$sparsity, "sparsity,", inf_sprout$time))
```

```{r eval=FALSE}
head(as.data.frame(inf_sprout$loadings))
```

```{r echo=FALSE}
kbl(head(inf_sprout$loadings), booktabs = T, caption = "Finite Lambda Loadings") %>%
  kable_styling(latex_options = c("striped", "repeat_header", "hold_position"))
```

We obtain the scree plot and the biplot using the same method.

```{r fig.align='center', fig.cap="Scree plot for lambda=inf case."}
pr.var <- (apply(inf_sprout$x, 2, sd))^2
pve <- pr.var / sum(pr.var)

par(mfrow=c(1,2))
plot(pve, 
     xlab="Sparse PC", 
     ylab="Proportion of Variance Explained", 
     ylim=c(0,1), 
     type="b")
plot(cumsum(pve), 
     xlab="Sparse PC", 
     ylab="Cumulative Proportion of Variance Explained", 
     ylim=c(0,1),
     type="b")
```

For data with high dimensionality, the biplot is still much harder to read with to a lower sparsity value.

```{r fig.align='center', fig.cap="Biplot for lambda=inf case. Observe that with lower sparsity in the loadings and high-dimensional data, the biplot becomes less readable."}
y_sub = apply(inf_sprout$loadings, 1, function(row) all(row != 0))
loadings = inf_sprout$loadings[y_sub, ]

par(mfrow=c(1,1))
biplot(inf_sprout$x, loadings, xlab="PC 1", ylab="PC 2")
```

\newpage

#### Python Example

Note that the Python package depends on numpy.

The following example accomplishes the same situation (down to the same randomly-generated data) but in Python.

```{python eval=FALSE}
import numpy as np
from sparsepca import spca

k = 4  # rank
p = 500  # dimensions
n = 1000  # sample size
lambda1 = 0.1 * np.ones((n, 1))
lambda2 = 1

np.random.seed(10)
z = np.random.normal(0, 1, size=(n, p))  # generate random normal 1000x500 matrix

fin_sprout = spca(z, lambda1, lambda2, k=k)
print(f"Finite: {fin_sprout['iter']} iterations with final value 
		{fin_sprout['f_manpg']}, sparsity {fin_sprout['sparsity']}, 
		timediff {fin_sprout['time']}.")

fin_sprout['loadings']

inf_sprout = spca(z, lambda1, np.inf, k=k)
print(f"Infinite: {inf_sprout['iter']} iterations with final value 
		{inf_sprout['f_manpg']}, sparsity {inf_sprout['sparsity']}, 
		timediff {inf_sprout['time']}.")

inf_sprout['loadings']
```

## References

Chen, S., Ma, S., Xue, L., and Zou, H. (2020) "An Alternating Manifold Proximal Gradient Method for Sparse Principal Component Analysis and Sparse Canonical Correlation Analysis" INFORMS Journal on Optimization 2:3, 192-208 <[doi:10.1287/ijoo.2019.0032](https://doi.org/10.1287%2Fijoo.2019.0032)>.

Zou, H., Hastie, T., & Tibshirani, R. (2006). Sparse principal component analysis. Journal of Computational and Graphical Statistics, 15(2), 265-286 <[doi:10.1198/106186006X113430](https://doi.org/10.1198%2F106186006X113430)>.

Zou, H., & Xue, L. (2018). A selective overview of sparse principal component analysis. Proceedings of the IEEE, 106(8), 1311-1320 <[doi:10.1109/JPROC.2018.2846588](https://doi.org/10.1109%2FJPROC.2018.2846588)>.

